---
title: "02. Revise Data"
author: "Perry Haaland"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(emmeans)
```

## Introduction

The purpose of the markdown document is to revise the existing data, to make it more suited for further analysis. In particular, one new variable is added (credits failed). All variables get R friendly names that will display better on graphics and generally require less typing.

This code produces a data set for common use. It is stored as an R objected in the data folder under the name `revised_df.RData`.


## Load the previously prepared data

We have all working from the following common data set up to this point.

```{r}
load("../RData/students_df.RData")
dim(students_df)
```

We recall the variable names.

```{r}
names(students_df)
```

## Variables for modeling

We are going to assign R friendly names, first.

* Retained = IS_RETAINED: 0 for did not return, 1 for returned 
* Residency = RESIDENCY: In-state, Out-of-state
* Class = CLASS_LEVEL_STUDENT: FR, SO, JR, and SR
* DoubleMajor = IS_DBL_MAJOR: 0 for No, 1 for Yes
* Enrollment = STU_ORIG_ENROLL_STATUS: New Student or Transfer Student 
* Credits = CREDIT_HOURS_EARNED
* CreditsFailed = CREDITS_FAILED: CREDIT_HOURS_ATT - CREDIT_HOURS_EARNED
* Program = MAJ_1_PROGRAM_DEGREE, abbreviation for program of the students first declared major
* CumulativeTerms = CUM_RESIDENT_TERMS_BOT, student's cumulative number of terms in residence
* TestCredits = TEST_CREDIT_ENTRY, test credits awarded at entry, for example, based on AP exams

## Setting contrasts for class level

As before, we start by making the class level an ordered factor based on student's year in school.

```{r}
class_levels <- sort(unique(students_df$CLASS_LEVEL_STUDENT))[c(1,3,2,4)]
mod_df <- students_df
mod_df$CLASS_LEVEL_STUDENT <-
  factor(students_df$CLASS_LEVEL_STUDENT, 
             levels = class_levels)
```

## Credits

We also add CREDITS_FAILED to the data. I like this variable a bit better than the credits earned because it seems to have a more easily grasped interpretation.


```{r}
mod_df <- mod_df %>%
  mutate(CREDITS_FAILED = CREDIT_HOURS_ATT - CREDIT_HOURS_EARNED)
```

## Some students dropped based on enrollment status

We also delete one very small group of students who don't easily fit in with the original enrollment status.

```{r}
mod_df <- mod_df %>%
  filter(STU_ORIG_ENROLL_STATUS != "Unclassified") %>%
  mutate(STU_ORIG_ENROLL_STATUS = 
           ifelse(STU_ORIG_ENROLL_STATUS == "New Student",
                  "New Student",
                  "Transfer Student"))
```


## Rename variables with R friendly names

Finally, for the purposes of this analysis and the ability to display the results, I rename many of the variables with R friendly and graphics suitable names.


```{r}
revised_df <- mod_df %>%
  select(Retained = IS_RETAINED,
         Class = CLASS_LEVEL_STUDENT,
         Residency = RESIDENCY,
         DoubleMajor = IS_DBL_MAJOR,
         Enrollment = STU_ORIG_ENROLL_STATUS,
         Credits = CREDIT_HOURS_EARNED,
         CreditsFailed = CREDITS_FAILED,
         Program = MAJ_1_PROGRAM_DEGREE,
         CumulativeTerms = CUM_RESIDENT_TERMS_BOT, 
         TestCredits = TEST_CREDIT_ENTRY)
```

There are 14,171 students in our final data set.

```{r}
nrow(revised_df)
```

## Checking on the variables

```{r}
revised_df %>%
  group_by(Class) %>%
  summarize(n = n(), .groups = "drop")
```


```{r}
revised_df %>%
  group_by(Residency) %>%
  summarize(n = n(), .groups = "drop")
```

```{r}
revised_df %>%
  group_by(DoubleMajor) %>%
  summarize(n = n(), .groups = "drop")
```


```{r}
revised_df %>%
  group_by(Enrollment) %>%
  summarize(n = n(), .groups = "drop")
```



```{r}
revised_df %>%
  group_by(Program) %>%
  summarize(n = n(), .groups = "drop")
```


```{r}
revised_df %>%
  group_by(Credits) %>%
  summarize(n = n(), .groups = "drop")
```


```{r}
revised_df %>%
  group_by(CreditsFailed) %>%
  summarize(n = n(), .groups = "drop")
```

```{r}
revised_df %>%
  group_by(CumulativeTerms) %>%
  summarize(n = n(), .groups = "drop")
```

```{r}
revised_df %>%
  group_by(TestCredits) %>%
  summarize(n = n(), .groups = "drop")
```
# Saving the data 

The revised file is saved in the data folder as `revised_df.RData`.

```{r}
save(revised_df, file = "../RData/revised_df.RData")
```

