---
title: "02_eda"
author: "Angel Huang"
date: "October 13, 2020"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2
```{r}
# Install EDA related libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(plyr)
library(dplyr) # for pipe
library(tidyr)
```

# 2.1 Load data
```{r}
load("INIT_TEST_FILE.Rda")
str(df) # list of each variable
```


```{r}
# Plot response variable
ggplot(data = df) +
  geom_bar(mapping = aes(x = IS_RETAINED_DESC))

```

From the bar plot, we see that there are about 1/4 students did not return and 3/4 students returned. Higher than my expectation, but maybe good for our model since there are enough cases with "not return". Note that there is still this imbalance of data that we might need to deal with later.

```{r}
# Plot continuous variables against response variable IS_RETAINED_DESC
p1 = ggplot(data = df, mapping = aes(x = IS_RETAINED_DESC, y = CREDIT_HOURS_EARNED)) +
  geom_boxplot()
p2 = ggplot(data = df, mapping = aes(x = IS_RETAINED_DESC, y = CREDIT_HOURS_ATT)) +
  geom_boxplot()

# How different is credit_hours_earned vs. attempted, with regression lines
p3 = ggplot(df, aes(x=CREDIT_HOURS_ATT, y=CREDIT_HOURS_EARNED)) + 
  geom_point()+
  geom_smooth(method=lm,fomular=CREDIT_HOURS_EARNED~CREDIT_HOURS_ATT)

grid.arrange(p1, p2, p3, nrow = 1) # arrange subplots (install gridExtra package)
```
It seems like students who returned next term have higher credit_hours_earned this semester. Plot against credit hours attempted has very similar result. After doing correlation, we found the two variables are highly correlated, maybe just 1 is needed in the model?


```{r}
# Plot categorical variables against response variable
ggplot(data = df) +
  geom_count(mapping = aes(x = IS_RETAINED_DESC, y = CLASS_LEVEL_STUDENT))

df_SR = df[df$IS_RETAINED == 0 & df$CLASS_LEVEL_STUDENT == 'SR',]
ggplot(data = df_SR) +
  geom_count(mapping = aes(x = STRM_DESC, y = FIRST_STRM_DESC))

count(df_SR, 'FIRST_STRM_DESC')
sum(3219+54+3+1)
```
Interestingly, most students who did not return next term are senior students, which might due to graduation!? We shouldn't count those cases for our purpose. To look into this, I plotted the first term of all senior students who is not retrained, mostly are from 2015 Fall, who actually graduate in Spring 2019. So most of the students in our negative cases are due to graduation, to be more specific, around 3219+54+3+1=3277 students are probably graduating.

```{r}
# Subsample dataset to exclude senior data
df_noSR = df[df$CLASS_LEVEL_STUDENT != 'SR',]
df_noSR_plot = df_noSR %>%
  group_by(MAJ_1_SCHOOL_NAME) %>%
  summarise(not_retained = mean(1-IS_RETAINED))
df_noSR_plot
```
```{r}
ggplot(df_noSR_plot,aes(x = MAJ_1_SCHOOL_NAME, y = not_retained*100)) +  
  geom_col() +
  coord_cartesian(ylim=c(0,10)) + # has to use coord_cartesian, which doesn't filter on data like ylim does
  theme(axis.text.x = element_text(angle =45))  # tilt x label

```
So the majors with the highest "drop" rate is medicine with 7% (this makes sense given the difficulty of the major), and then arts and sciences with 5%, and then Dentistry, Nursing, Pharmacy, and business, each around 4%. 
