---
title: "week5 teamwork"
author: "Zhaoqi Liu"
date: "12/7/2020"
output: word_document
---
```{r}
load("~/Downloads/STOR 765 consulting/STOR765_FALL2020_TeamDataScience/RData/revised_df.RData")
```

```{r,message=FALSE}
library(ggplot2)
library(dplyr)
```

```{r}
df<-revised_df
df$Residency<-as.factor(revised_df$Residency)
df$DoubleMajor<-as.factor(revised_df$DoubleMajor)
df$Enrollment<-as.factor(revised_df$Enrollment)
df$Program<-as.factor(revised_df$Program)
```

```{r}
#if include program's interaction, the algorithm did not converge 
fit_all <- glm(Retained ~ (Class +
                                Residency +
                                DoubleMajor + 
                                Enrollment + 
                                Credits +
                                CreditsFailed+
                                TestCredits)^2+Program, 
            data = df,
            family = binomial(link = "logit"))

```

```{r}
#summary(fit_all)
```


```{r}
anova(fit_all, test = "LR")
```

###drop some interaction terms 
choose interaction term based on the anova test above and test if the reduced model has lack of fit 
```{r}
fit_reduced <- glm(Retained ~ Class + Residency + DoubleMajor + Enrollment + 
                              Credits + CreditsFailed+TestCredits + Program +
                            DoubleMajor*(Class+CreditsFailed) + 
                            Credits*(Class+DoubleMajor+Enrollment+
                                       CreditsFailed+TestCredits) +
                            Class*Enrollment,
                            
            data = df,
            family = binomial(link = "logit"))
```

```{r}
summary(fit_reduced)
anova(fit_reduced, test="LR")
```

Compare reduced model and full model: there is no lack of fit. 
The reduced model contains less interaction terms: doublemajor with class_level, credits and credits_failed. Credits with all the other features, and Class*Enrollment 
```{r}
anova(fit_reduced,fit_all,test="LR")
```

Add prediced retation rate from reduced model to df.  
```{r}
pred<-predict(fit_reduced,type="response")
df$pred<-pred
```

Create new columns for categorical credits and creditsfailed 
Since the minimum course load for undergraduates for a single semester is 12 academic credit hours. For students who have less than 12 credits for a semester, they need the permission from their dean. So we classify credits(actual earned) to two categories: below 12 and above 12. We also classify creditsfailed into sequence with  an interval 3 since most classes are 3 credits.
```{r}
cat_creditfail<-cut(revised_df$CreditsFailed,breaks = seq(-3,18,3),include.lowest = TRUE)
table(cat_creditfail)

cat_credits<-ifelse(revised_df$Credits<12,"less than 12 credits","greater than or equal to 12 credits" )
#levels(cat_credits)
cat_credits<-factor(cat_credits, levels = c("less than 12 credits","greater than or equal to 12 credits"))
cat_credit2<-cut(revised_df$Credits,breaks = c(seq(0,18,3),22),include.lowest = TRUE, right=FALSE)
table(cat_credit2)
```

```{r}
df$cat_creditfail<-cat_creditfail
df$cat_credit<-cat_credits
df$cat_credit2<-cat_credit2
```

```{r}
#return the position and text lable 
getN <- function(x){
  return(c(y = 0, label = length(x)))
}
getMean <- function(x){
  return(c(y = mean(x), label = round(mean(x),digit=2)))
}
```

###Double Major with Credit Fails 
add the total number of each category to the plot 
```{r}
df%>%ggplot(aes(x=cat_creditfail, y=pred,fill=DoubleMajor))+
  geom_boxplot(outlier.size = 0.1) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "Predicted Retation Rate with Double Major and Failed Credits")+
  xlab("Credits Failed")+
  ylab("Predicted Retation Rate")
```

For students who did't fail any credits, whether they have a second major doesn't affect the predicted rate much, but students with one major have more outliers. 

```{r}
#don't know how to change it 
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}
df1<-df
df1%>%
  group_by(CreditsFailed,DoubleMajor) %>%
  mutate(outlier = ifelse(is_outlier(pred), pred, as.numeric(NA)))
df1%>%ggplot(aes(x = cat_creditfail, y = pred, fill=DoubleMajor)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(aes(x = cat_creditfail,y = pred, fill=DoubleMajor, color = factor(Retained),shape ="."), na.rm = TRUE)
```

plot the retation rate with double major, credit fails and the true retation result 
```{r}
p<-df%>%ggplot(aes(x=cat_creditfail, y=pred))+
  geom_point(aes(color=factor(Retained))) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "True Retation with Double Major and Failed Credits")+
  scale_color_discrete(name = "Retained", labels =c("No","Yes"))+
  xlab("Credits Failed")+
  ylab("Predicted Retation Rate")
p+facet_grid(col = vars(DoubleMajor), labeller = as_labeller(c(`0`="Single Major",`1`="Double Major")))
```


plot the predicted retation rate with double major and credits earned
```{r}
df%>%ggplot(aes(x=cat_credit2, y=pred,fill=DoubleMajor))+
  geom_boxplot(outlier.size = 0.1) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "Predicted Retation Rate with Double Major and Earned Credits")+
  xlab("Credits ")+
  ylab("Predicted Retation Rate")
```


plot the predicted retation rate with double major, credits earned and the true retation result 
```{r}
p<-df%>%ggplot(aes(x=cat_credit2, y=pred))+
  geom_point(aes(color=factor(Retained))) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "True Retation with Double Major and Earned Credits")+
  scale_color_discrete(name = "Retained", labels =c("No","Yes"))+
  xlab("Credits Earned")+
  ylab("Predicted Retation Rate")
p+facet_grid(col = vars(DoubleMajor), labeller = as_labeller(c(`0`="Single Major",`1`="Double Major")))
```


If treat credits earned as a two-category variable: <12 and >=12
```{r}
df%>%ggplot(aes(x=cat_credit, y=pred,fill=DoubleMajor))+
  geom_boxplot(outlier.size = 0.1) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "Double Major with two-category Earned Credits")+
  xlab("Credits ")+
  ylab("Predicted Retation Rate")
```

For students whose Credits <12, whether they have a second major plays an important role in determining whether they will come back. 

```{r}
p2<-df%>%ggplot(aes(x=cat_creditfail, y=pred,fill=DoubleMajor))+
  geom_boxplot(outlier.size = 0.1) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "Predicted Retation Rate with Double Major, Failed Credits and two-category Credits")+
  xlab("Credits Failed")+
  ylab("Predicted Retation Rate")
p2+facet_grid(~cat_credit)
```

For students who have earned less than 12 credits in this semester, which means undergraduate students cannot be regard as a full-time students at UNC, double major affects the retation. Those students who has a second major are more likely to return. For students who have earned more than 12 credits, double major doesn't affect much on the retation rate prediction. 


```{r}
p3<-df%>%ggplot(aes(x=Program, y=pred,fill=DoubleMajor))+
  geom_boxplot(outlier.size = 0.1) + 
  stat_summary(fun.data = getN, geom = "text",position = position_dodge(width = 0.75), size=2.5)+
  stat_summary(fun.data = getMean, geom="text",position = position_dodge(width = 0.75),size = 2.5)+
  labs(title = "Predicted Retation Rate with Double Major, Program and Earned Credits")+
  xlab("Program")+
  ylab("Predicted Retation Rate")
p3+facet_grid(row=vars(cat_credit))
```

ASBCH seems to be program related with natural science and mathematics. 
Considering the Programs, we see that students in some programs do not have a second major. The number of students in BA programs who obtain less than 12 credits in this semester, is larger than that of BS program. 
For students who obtain less than 12 credits this semester, students with a second major are more likely to return. 

```{r}
df%>%filter(cat_credit=="less than 12 credits" )%>%
  group_by(Retained,Class,DoubleMajor)%>%
  summarise(n = n())
```

We can see that for students has less than 12 credits, the number of students with a second major is less than the number of students with single major for all class level. 
